Wait-free Queues with Polygarithmic Step Complexity

### 2022-05-18 article/psuedocode.tex
apply may 13 notes

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

@@

\Function{boolean}{Refresh()}{}
\State \tt{s= size} \label{readSize}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

@@

\begin{definition}[Subblock]
Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} or  is a subblock of a block in it.
\end{definition}

For simplicity we say block \nf{b} is propagated to node \nf{n} or set \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.

@@

\begin{definition}
  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$. Block \nf{b} in \nf{n.blocks} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

@@

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{b} is the block returned by \nf{n.CreateBlock(h, x)} invoked at time $t$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
\end{lemma}
\begin{proof}
 WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock. Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s blocks after line~\ref{secondRefresh}.
\end{corollary}
%\begin{proof}
%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35. By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}. \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well. First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively. If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails. It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
%\end{proof}

@@

\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
Each block has at most $p$ subblocks.
\end{lemma}
%\begin{proof}
%  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
%\end{proof}

@@

\begin{itemize}
\item We call operations before $op$ in the sequence of operations $S$, prefix of the $op$.
  \item $E(n,i)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
  \item $D(n,i)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
\item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on deqeueus.}

@@

\begin{lemma}[Get corretness] \label{get}
\nf{n.GetENQ(b,i)} returns the $i$th Enqueue in $E(n,b)$.
\end{lemma}
%\begin{proof}
%It is obvious that \nf{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$. To find the \nf{i}th enqueue in block \nf{b} of an internal node \nf{n} in line 87 it is decided that it resides in the left child or the right child. This decision is made by Definition of $E(n,b)$. After that Lines 88, 92 search the proper subblocks of \nf{b}. From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \nf{prevBlock} and \nf{lastBlock} block of the \nf{CreateBlock()}.
%\end{proof}

\begin{lemma}[Index correctness]
 \nf{n.Index(b,i)} returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
\end{lemma}
%\begin{proof}
%  \nf{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \nf{n} in \nf{n.parent} and then computes the order in $D(n.parent, superblock)$. Then calls \nf{index()} on \nf{n.parent} recursively. It is easy to see why the second is correct. Correctness of computing superblock comes from Lemma \ref{superBlock}.
%\end{proof}

\begin{lemma}[Computing Queue's Head] \label{computeHead}
  Let S be the state of an empty queue if the operations in prefix of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns the index of the enqueue that is the head in S. If the  queue is empty in S it returns \nf{<-1,-->}.
  \end{lemma}

\begin{lemma}[Validity of \nf{head}]\label{head}
If block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} is going to remain b (not overwritten in future).
\end{lemma}
%\begin{proof}
%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read. If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}. If a process dies before line 51, another process will increment \nf{head} in line 54.
%\end{proof}

\begin{lemma}[help]\label{help}
After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone} are done.
\end{lemma}

>>>>

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

@@

\Function{boolean}{Refresh()}{}
\State \tt{s= size} \label{readSize}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, s, s+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, s, s+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?

@@

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
\end{definition}

For simplicity we say block \nf{b} is propagated to node \nf{n} or to a set of blocks \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.

@@

\begin{definition}
  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$. Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

\begin{observation}[Validity of \nf{head}]\label{head}
Once block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
\end{observation}

\begin{lemma}
  Each block is not duplicated.
\end{lemma}

@@

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $b$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
\end{lemma}
\begin{proof}
 WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the established operations at time $t$ by definition of subblock. Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

@@

\begin{corollary}[Propagate Step] \label{doublyRefresh}
All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
\end{corollary}
%\begin{proof}
%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35. By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}. \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well. First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively. If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails. It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
%\end{proof}

@@

\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
Each block has at most $p$ direct subblocks.
\end{lemma}
%\begin{proof}
%  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
%\end{proof}

@@

\begin{itemize}
\item We call operations strictly before $op$ in the sequence of operations $S$, prefix of the $op$.
  \item $E(n,i)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
  \item $D(n,i)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
\item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on dequeues.}

\begin{lemma}[Get correctness] \label{get}
\nf{n.GetENQ(b,i)} returns the $i$th Enqueue in $E(n,b)$.
\end{lemma}
%\begin{proof}
%It is obvious that \nf{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$. To find the \nf{i}th enqueue in block \nf{b} of an internal node \nf{n} in line 87 it is decided that it resides in the left child or the right child. This decision is made by Definition of $E(n,b)$. After that Lines 88, 92 search the proper subblocks of \nf{b}. From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \nf{prevBlock} and \nf{lastBlock} block of the \nf{CreateBlock()}.
%\end{proof}

\begin{lemma}[help]\label{help}
After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
\end{lemma}

\begin{lemma}[Index correctness]
 \nf{n.Index(b,i)} returns the rank in $D(root)$ of the $i$th Dequeue in $D(n,b)$.
\end{lemma}
%\begin{proof}
%  \nf{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \nf{n} in \nf{n.parent} and then computes the order in $D(n.parent, superblock)$. Then calls \nf{index()} on \nf{n.parent} recursively. It is easy to see why the second is correct. Correctness of computing superblock comes from Lemma \ref{superBlock}.
%\end{proof}

\begin{lemma}[Computing Queue's Head] \label{computeHead}
  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it. \nf{FindResponse()} returns the index in $E(root,b)$ of the enqueue that is the head in S. If the  queue is empty in S it returns \nf{<-1,-->}.
  \end{lemma}

%\begin{proof}
%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read. If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}. If a process dies before line 51, another process will increment \nf{head} in line 54.
%\end{proof}

@@
